---
title: "01-00. Run WGCNA pipeline"
author: "Rachel Smith"
format: html
editor: visual
---

Run WGCNA pipeline on case-control data:

\(1\) Select soft-thresholding power

\(2\) Calculate adjacency matrix & TOM matrix

\(3\) Cluster and assign modules

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
base_dir <- "~/Documents/PhD/projects/sgacc_wgcna_grcca/"
```

## Setup and Load Data

```{r load_data, message=FALSE, warning=FALSE}
source(paste0(base_dir, "scripts/load_genes.R"))
```

## (1) Select soft thresholding power

```{r sft}
## Convert normalized & regressed counts to transposed matrix
m_vsd_regress <- df_vsd_regress %>% column_to_rownames("sample")

## Choose a set of soft-thresholding powers to test
powers <- 1:16

## Call the network topology analysis function
sft <- pickSoftThreshold(m_vsd_regress, powerVector = powers, verbose = 5)

## Convert results to a tibble
df_sft <- sft$fitIndices %>%
    as_tibble() %>%
    clean_names()

## Save
prefix <- "08Mar2024_GENES_qSVAgeSexRaceGC_SIGNED" # Defining prefix if not already defined
save(df_sft, file = paste0(objects_dir, prefix, "_SFT.RDS")) # for downstream analyses
```

## (2) Calculate adjacency matrix, TOM, clustering

```{r adjacency}
doParallel::registerDoParallel()

## Select sft
soft_power <- 3

# Calculate co-expression similarity/adjacency
set.seed(20240306)
adjacency <- adjacency(m_vsd_regress, type = "signed hybrid", power = soft_power)

## Compute topological overlap matrix (TOM)
set.seed(20240306)
TOM <- TOMsimilarity(adjacency, TOMType = "signed")
diss_TOM <- 1 - TOM

## Run hierarchical clustering on TOM
gene_tree <- hclust(as.dist(diss_TOM))
```

## (3) Module assignment

```{r module_assignment}
## Assign genes to modules across minimum module size & cut height parameter space
df_combos <- expand_grid(
    min_size = c(30, 35, 40), # c(30, 40, 50),
    cut_height = seq(from = 0.95, to = 0.98, by = 0.003)
)

set.seed(20240306)
df_modules <- map2_dfr(
    .x = df_combos$min_size,
    .y = df_combos$cut_height,
    .f = ~ tibble(
        sft = soft_power,
        min_size = .x,
        cut_height = .y,
        ensembl_gene_id = colnames(m_vsd_regress),
        module = cutreeDynamic(
            dendro = gene_tree,
            distM = diss_TOM,
            pamRespectsDendro = FALSE,
            minClusterSize = .x,
            cutHeight = .y
        )
    )
) %>%
    mutate(module = factor(module, levels = 0:(max(module))))

df_modules %>%
    dplyr::count(min_size, cut_height, module) %>%
    print(n = nrow(.))

## Plot module sizes
df_module_sizes <- df_modules %>%
    group_by(sft, min_size, cut_height) %>%
    dplyr::count(module)

df_module_sizes %>%
    filter(cut_height != 1) %>%
    ggplot(aes(x = module, y = n)) +
    geom_col(width = 0.5) +
    geom_point(shape = 21, size = 0.5) +
    geom_vline(aes(xintercept = 21), lty = 2, color = "red") +
    scale_x_discrete(breaks = seq(0, 50, 10)) +
    facet_grid(min_size ~ cut_height) +
    ggtitle(paste0("SFT", soft_power, " module assignments"))

# SELECTION: SFT = 3, minimum module size = 40, cut_height = 0.98
```

## Assign module colors

```{r assign_colors}
## Assign colors to modules (using a different color palette than the WGCNA default)

## Select palette
igv_palette <- paletteer_d("ggsci::default_igv")

## Remove light colors and gray
igv_palette <- igv_palette[!(igv_palette %in% c("#F0E685FF", "#CDDEB7FF", "#5A655EFF", "#A9A9A9FF"))] # n = 47

## Assign to modules in the tibble
df_colors <- tibble(
    module = 0:max((df_modules$module) %>% 
                       as.character() %>% 
                       as.numeric()) %>% 
        as.factor()
) %>%
    mutate(
        color = ifelse(module == 0, "#A9A9A9FF", igv_palette)
    )

df_modules <- df_modules %>%
    left_join(df_colors, by = join_by(module)) %>%
    arrange(sft, min_size, cut_height, module)
```

## Save final module assignments

```{r save}
save(df_modules, file = paste0(objects_dir, prefix, "_SIGNED_SFT", soft_power, "_MODS.RDS"))
```

## Plots

### Figure 1C \| WGCNA dendrogram

```{r Fig1c.plot, fig.height = 4, fig.width = 8}
# Pull colors for the selected parameters (min_size = 40, cut_height = 0.98)
selected_colors <- df_modules %>% 
    filter(min_size == 40 & cut_height == 0.98) %>% 
    pull(color)

plotDendroAndColors(gene_tree, selected_colors,
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "C | WGCNA cluster dendrogram")

# Save
pdf(paste0(figures_dir, "C.WGCNA_dendrogram.pdf"), width = 8, height = 4)
plotDendroAndColors(gene_tree, selected_colors,
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "C | WGCNA cluster dendrogram")
dev.off()
```

### Figure S6A \| Soft-thresholding power selection

```{r FigS6a.plot, fig.width = 5, fig.height = 2}
## Model fit of each soft thresholding power
p1 <- df_sft %>% 
    ggplot(aes(x = power, y = -sign(slope)*sft_r_sq)) +
    geom_point(aes(color = power == 3), size = 1) +
    geom_hline(yintercept = 0.9, lty = 2, color = "maroon") +
    scale_color_manual(values = c("TRUE" = "maroon", "FALSE" = "black"), guide = "none") +
    scale_x_continuous(breaks = seq(2, 16, by = 2)) +
    labs(
        x = "Soft threshold (power)",
        y = "Scale free topology model fit \n(signed R^2)"
    ) +
    theme(
        panel.grid.major.x = element_line(color = "darkgrey", size = 0.2)
    )

## Median connectivity as a function of the soft-thresholding power 
## (want to remain reasonably high (in the 100s) or above)
p2 <- df_sft %>% 
    ggplot(aes(x = power, y = median_k)) +
    geom_point(aes(color = power == 3), size = 1) +
    #geom_text(aes(label = power), size = 3) +
    geom_hline(yintercept = 100, lty = 2, color = "maroon") +
    scale_color_manual(values = c("TRUE" = "maroon", "FALSE" = "black"), guide = "none") +
    scale_x_continuous(breaks = seq(2, 16, by = 2)) +
    labs(
        x = "Soft threshold (power)",
        y = "Median connectivity"
    ) +
    theme(
        panel.grid.major.x = element_line(color = "darkgrey", size = 0.2)#,
    )

## Combine and save
p1 + p2 + plot_annotation(title = "A | Soft-thresholding power fit to data")
map(
  .x = c(".png", ".pdf"),
  .f = ~ ggsave(paste0(supp_figures_dir, "S6A.SFT_fit", .x), 
                width = 5, height = 2)
)
```

### Figure S6B \| Module stability across parameters

```{r FigS6b.plot, fig.width = 5, fig.height = 6.5}
df_modules %>%
    group_by(sft, min_size, cut_height) %>%
    dplyr::count(module, color) %>% 
    #filter(cut_height != 1) %>%
    
    ggplot(aes(x = module)) +
    geom_segment(aes(y = 0, yend = log10(n), color = color)) +
    geom_point(aes(y = log10(n), fill = color), shape = 21, size = 0.75) +
    geom_vline(aes(xintercept = 21), lty = 2, color = "maroon") +
    scale_fill_identity() +
    scale_color_identity() +
    scale_x_discrete(breaks = seq(0, 50, 10)) +
    facet_grid(cut_height ~ min_size) +
    labs(y = "log10(N genes)", 
         title = "B | Module assignments across WGCNA parameters"
    ) +
    theme(
        strip.text = element_text(size = 9),
        axis.title = element_text(size = 8),
        axis.text = element_text(size = 8)
    )
map(
  .x = c(".png", ".pdf"),
  .f = ~ ggsave(paste0(supp_figures_dir, "S6B.module_sizes", .x), 
                width = 5, height = 6.5)
)
```
