---
title: "01_characterize-module-enrichment"
author: "Rachel Smith"
format: html
editor: visual
---

# Characterize the biological enrichments of the assigned modules

Run GO and cell-type enrichment on WGCNA modules (gene-level). Goal is to understand what each module represents biologically. Includes GO (+ topic modeling), cell type, and developmental trajectory.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
base_dir <- "~/Documents/PhD/projects/sgacc_wgcna_grcca/"
objects_dir <- paste0(base_dir, "objects/")
analysis_objects_dir <- paste0(base_dir, "outputs/objects/")
```

## Setup

```{r load_data}
source(paste0(base_dir, "code/analysis/load_WGCNA_res.R")) ## selects final module set (df_modules_filt)
library(clusterProfiler)
library(tidytext)
library(stm)
library(tidyverse)
```

## Run GO on each WGCNA module

Run GO pathway overrepresentation analysis on each module using enrichGO

```{r run_go}
modules <- df_modules_filt %>%
    pull(module) %>%
    unique()

doParallel::registerDoParallel()

df_mods_go <- map_dfr(
    .x = modules,
    .f = ~ enrichGO(
        gene = df_modules_filt %>% dplyr::filter(module == .x) %>% pull(ensembl_gene_id),
        OrgDb = "org.Hs.eg.db",
        universe = ensembl_gene_ids,
        keyType = "ENSEMBL",
        ont = "ALL",
        pAdjustMethod = "BH",
        pvalueCutoff = 1,
        qvalueCutoff = 1,
        readable = TRUE
    ) %>%
        as_tibble() %>%
        mutate(module = .x, .before = 1)
)


## Save results
prefix <- "08Mar2024_GENES_qSVAgeSexRaceGC" # Assuming prefix
soft_power <- 3 # Assuming soft_power
save(df_mods_go,
    file = paste0(objects_dir, prefix, "_SIGNED_SFT", soft_power, "_GO_RES.RDS")
)
```

## Run topic modeling on GO results for each module

Goal: understand what biological processes each module represents using the semantic similarity across GO term descriptions

```{r topic_modeling}
## Unnest tokens to get GO terms within each module (group)
df_go_tokens <- df_mods_go %>%
    dplyr::filter(pvalue < 0.05) %>%
    dplyr::filter(!is.na(Description)) %>%
    group_by(module) %>%
    unnest_tokens(word, Description)

## Create a sparse matrix using only meaningful words that appear at least twice within module
vague_terms <- c(
    "positive", "negative", "regulation", "response", "protein", "activity",
    "pathway", "process", "involved", "signaling"
)
go_sparse <- df_go_tokens %>%
    dplyr::count(module, word) %>%
    mutate(module = as.numeric(as.character(module %>% str_remove("geneM")))) %>%
    anti_join(stop_words) %>%
    dplyr::filter(n > 2 & !(word %in% vague_terms)) %>%
    cast_sparse(module, word, n)

dim(go_sparse)

## Fit topic model
set.seed(456)
n_topics <- 5
topic_model <- stm(go_sparse, K = n_topics, verbose = FALSE)


## Save the topic model and sparse matrix for plotting
save(go_sparse, topic_model,
    file = paste0(analysis_objects_dir, "topic_modeling_res.Rdata")
)
```

## Cell-type enrichment

Run hypergeometric test of each module with each Lake cell type

```{r cell_type}
df_mods_cell_type_hypergeometric <- expand_grid(
    cell_type = names(cell_types),
    module = modules
) %>%
    mutate(
        hypergeometric_res = map2(
            .x = cell_type,
            .y = module,
            .f = ~ f_module_hypergeometric(
                gene_list = cell_types[[.x]],
                gene_list_name = .x,
                wgcna_module = .y,
                gene_universe = ensembl_gene_ids,
                module_data = df_modules_filt
            ) %>%
                dplyr::select(-module)
        )
    ) %>%
    unnest(cols = c(hypergeometric_res)) %>%
    mutate(p_adj = p.adjust(p_value, method = "fdr"))

## Save for plotting
save(df_mods_cell_type_hypergeometric,
    file = paste0(analysis_objects_dir, "cell_type_res.RDS")
)
```

## Approximate developmental trajectory of each module

```{r development}
load(paste0(base_dir, "objects/psychencode_development_expr_data.Rdata")) # df_psychencode_metadata, df_psychencode_expr

## Identify frontal cortex samples
df_sample_window <- df_psychencode_metadata %>%
    dplyr::filter(region_broad == "frontal_cortex")
frontal_cortex_samples <- df_sample_window %>%
    pull(id)

## Filter expression data for samples and genes of interest
df_psychencode_expr_filt <- df_psychencode_expr %>%
    dplyr::filter(GENE %in% ensembl_gene_ids) %>%
    dplyr::select(GENE, all_of(frontal_cortex_samples)) %>%
    dplyr::rename("ensembl_gene_id" = "GENE")

## Combine expresion data and module data
df_expr_mod <- df_psychencode_expr_filt %>%
    pivot_longer(starts_with("HS"), names_to = "id", values_to = "expression") %>%
    left_join(df_sample_window) %>%
    left_join(df_modules_filt) %>%
    dplyr::select(ensembl_gene_id, module, id, window, expression)

## Find median expression value of each sample for each module
df_expr_mod_median <- df_expr_mod %>%
    group_by(module, id, window) %>%
    summarise(median_expr = median(expression))

## Save for plotting
save(df_expr_mod_median,
    file = paste0(objects_dir, "developmental_trajectory.RDS")
)
```
