---
title: "06c_SENS-mu"
author: "Rachel Smith"
format: html
editor: visual
---

# Run GRCCA across mu values to demonstrate sensitivity of the main LV

Demonstrate sensitivity of the main LV to the group penalty (mu ∈ {0.001 0.01 0.5 0.9 0.999}; quasi-logarithmic scale).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
base_dir <- "~/Documents/PhD/projects/sgacc_wgcna_grcca/"
scripts_dir <- paste0(base_dir, "scripts/")
figures_dir <- paste0(base_dir, "outputs/figures/response_to_reviewers/round2/")

# Create figures directory if it doesn't exist
if(!dir.exists(figures_dir)) dir.create(figures_dir, recursive = TRUE)
```

## Setup

```{r load_data}
source(paste0(base_dir, "scripts/load_WGCNA_res.R"))
library(ggsci)
library(ggpubr)
library(ComplexUpset)
```

## Load mu sensitivity results

```{r load_mu_results}
## Identify analysis directories for separate diagnosis runs
project_dir <- paste0("08Mar2024_GENES_qSVAgeSexRaceGC_sft3_minSize40_cutHeight0.98_ControlBDMDDSCZ_8MCA_regressBrainWeight_WITHGRAY/")
cca_dir <- paste0(base_dir, "RCCA_toolkit/GENES/", project_dir)


## Define variables altered for analysis
analysis_type <- "grcca" # rcca
analysis_framework <- "permutation" # holdout
VARx <- "0.1_1" # variance explained search grid (or set to specific value)
lambda <- "Nfeat"
mu <- c(0.001, 0.01, 0.5, 0.9, 0.999) # 0.1 is the original value
include_cmat <- "noCmat"


## Define analysis directory (variables altered in analysis with same X.mat and Y.mat)
analysis_dirs <- paste0(
    analysis_type, "_",
    analysis_framework, "_",
    "muSENS", "_", # turned on for mu sensitivity analysis; don't normally have this flag in
    "VARx", VARx, "_",
    "L2x", lambda, "_",
    "groupL2x", mu, "_",
    include_cmat
)


## Load GRCCA results for each analysis directory (across mu)
l_grcca_res <- 
    map(
        .x = analysis_dirs,
        .f = ~ f_load_cca_res(
            cca_directory = cca_dir,
            analysis_directory = .x,
            level = "gene",
            rename_covariates = TRUE
        )
    )
names(l_grcca_res) <- as.character(mu)
```

## Combine results into data frames for plotting

```{r combine_results}
## Load original GRCCA results
load(paste0(base_dir, "outputs/objects/GRCCA_results.Rdata")) # df_lvs, df_results, df_y_res, df_x_res, df_rx_expr_cor


## Model results
df_results_muSENS <- map(
    .x = l_grcca_res,
    .f = ~ .x$model_results
) %>%
    list_rbind(names_to = "mu") %>%
    bind_rows(df_results %>% mutate(mu = "0.1"))


## Y results
df_y_res_muSENS <- map(
    .x = l_grcca_res,
    .f = ~ .x$y_res %>% 
        mutate_at(vars(c(weight, z_score, pearsons_r)), ~ -.x) # negate to align with original model
) %>%
    list_rbind(names_to = "mu") %>%
    bind_rows(df_y_res %>% mutate(mu = "0.1"))


## X results
df_x_res_muSENS <- map(
    .x = l_grcca_res,
    .f = ~ .x$x_res %>% 
        mutate_at(vars(c(weight, z_score, pearsons_r)), ~ -.x) # negate to align with original model
) %>%
    list_rbind(names_to = "mu") %>%
    left_join(df_ensembl_to_symbol) %>% # Add gene & module info
    left_join(df_modules_filt %>% dplyr::select(-c(mod_set, color)), by = join_by(ensembl_gene_id)) %>%
    bind_rows(df_x_res %>% mutate(mu = "0.1"))
```

## Figure R3C2A | Mu sensitivity model comparison

```{r FigR3C2A.plot, fig.width=6.5, fig.height=2.1}
# Plot
df_results_muSENS %>% 
    
    # format
    group_by(mu) %>% 
    mutate(best_mod = ifelse(pval == min(pval), "black", "transparent")) %>%
    mutate(best_mod = ifelse(mu == 0.9 & set == 4, 0, best_mod)) %>%
    ungroup %>% 
    
    # layout
    ggplot(aes(x = factor(varx), y = -log10(pval))) +
    geom_hline(yintercept = -log10(0.005), lty = 2) +
    geom_segment(aes(y = 0, yend = -log10(pval), color = mu),
                 linewidth = 0.5, position = position_dodge(width = 0.6)) +
    geom_point(aes(size = correl, fill = mu, color = I(best_mod)),
               shape = 21, position = position_dodge(width = 0.6)) +
    
    # plot aesthetics
    scale_fill_nejm() +
    scale_color_nejm() +
    scale_size_continuous(range = c(0.5, 3), name = "X-Y LV cor") +
    labs(x = "X variance explained", y = "-log10(model P value)",
         title = "A | Mu sensitivity model comparison") +
    theme(
        legend.title = element_text(margin = margin(t = -5)),
        legend.margin = margin(l = -10)
    )


## Save plot
map(
    .x = c(".png", ".pdf"),
    .f = ~ ggsave(paste0(figures_dir, "R3C2A.mu_sensitivity_model_results", .x), width = 6.5, height = 2.1)
)
```

## Figure R3C2B | Mu sensitivity covariate results

```{r FigR3C2B.plot, fig.width=6.5, fig.height=2.1}
# Set covariate order for plotting
covariate_order <- c(
    "SCZ", "BD", "MDD",
    "Composite\ncase",
    paste0("Dim ", 1:8)
)


# Plot
df_y_res_muSENS %>%
    
    # format
    mutate(
        type = ifelse(str_detect(covariate, "Dim"), "tox", "main"),
        significant = ifelse(p_adj < 0.05 & abs(z_score) > 2, "black", "transparent"),
        covariate = factor(covariate, levels = covariate_order)
    ) %>% 
    arrange(abs(pearsons_r)) %>% 
    
    # layout
    ggplot(aes(x = covariate, y = pearsons_r)) +
    geom_segment(aes(y = 0, yend = pearsons_r, color = mu),
                 position = position_dodge(width = 0.7), linewidth = 0.5) +
    geom_point(aes(size = -log10(p_adj + 10^-100), fill = mu, color = I(significant)), 
               position = position_dodge(width = 0.7), shape = 21) +
    geom_hline(yintercept = 0, color = "gray", linewidth = 0.5) +
    
    # Facet by covariate type
    facet_wrap(vars(type), scales = "free_x", nrow = 1) +
    force_panelsizes(cols = c(3, 8)) +
    
    # Plot aesthetics
    scale_fill_nejm(guide = "none") +
    scale_color_nejm(guide = "none") +
    scale_size_continuous(range = c(0.5, 3), name = "-log10(FDR)") +
    #coord_cartesian(clip = "off") +
    labs(x = NULL, y = "Structure correlation (ry)",
         title = "B | Covariate results") +
    theme(
        strip.text = element_blank(),
        panel.spacing = unit(0.25, "cm"),
        legend.position = c(0.75, 0.75),
        legend.key.size = unit(0.25, "cm"),
        legend.title = element_text(margin = margin(t = -2))
        #legend.text = element_text(size = 5)
    )


## Save
map(
    .x = c(".png", ".pdf"),
    .f = ~ ggsave(paste0(figures_dir, "R3C2B.mu_sensitivity_covariate_results", .x), width = 6.5, height = 2.1)
)
```

## Figure R3C2C | Gene structure correlations across mu values

```{r FigR3C2C.plot, fig.width=6.5, fig.height=1.75}
## Format to correlate x res at each mu sensitivity value with actual x res
df_x_res_muSENS_plot <- df_x_res_muSENS %>% 
    dplyr::select(mu, ensembl_gene_id, pearsons_r) %>% 
    filter(mu != "0.1") %>% 
    left_join(
        df_x_res %>% 
            dplyr::select(ensembl_gene_id, pearsons_r) %>% 
            dplyr::rename("actual_r" = "pearsons_r"),
        by = join_by(ensembl_gene_id)
    )

## Plot
df_x_res_muSENS_plot %>% 
    
    # layout
    ggplot(aes(x = pearsons_r, y = actual_r)) +
    geom_abline(color = "gray") +
    geom_point(size = 0.25) +
    geom_smooth(method = "lm", color = "maroon", linewidth = 0.5) +
    stat_cor(aes(label = after_stat(r.label)), size = 3) +
    facet_wrap(vars(mu), nrow = 1) +
    
    # aesthetics
    scale_y_continuous(limits = c(-1, 1), breaks = c(-1, 0, 1)) +
    scale_x_continuous(limits = c(-1, 1), breaks = c(-1, 0, 1)) +
    labs(x = "Mu sensitivity rx", y = "Original rx (mu = 0.1)",
         title = "C | Gene structure correlations across mu values") +
    theme_bw() +
    theme(
        plot.title = element_text(face = "bold", size = 11),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 9),
        strip.text = element_text(margin = margin(t = 1, b = 1))
    )


## Save
map(
    .x = c(".png", ".pdf"),
    .f = ~ ggsave(paste0(figures_dir, "R3C2C.mu_sensitivity_gene_results", .x), width = 6.5, height = 1.75)
)
```

## Figure R3C2D | Gene-level overlaps (after threshold for significance)

```{r FigR3C2D.plot, fig.width=6.5, fig.height=3}
## Define mu values to plot across
df_xres_muSENS_upset_prep <- df_x_res_muSENS %>%
    mutate(significant = ifelse(significant == 1, TRUE, FALSE)) %>% 
    arrange(mu) %>% 
    pivot_wider(id_cols = ensembl_gene_id, names_from = mu, values_from = significant) %>% 
    filter(if_any(starts_with("0"), ~ . == TRUE))

mu_sets <- colnames(df_xres_muSENS_upset_prep %>% dplyr::select(-ensembl_gene_id))


## Upset plot
p_r3c2d <- upset(
    
    df_xres_muSENS_upset_prep, 
    mu_sets,
    stripes = "white", # remove gray strips from intersections matrix
    wrap = TRUE, # wrap plot to put title on the top
    sort_sets = "FALSE",
    
    # Simplify base annotations and control text size
    base_annotations = list(
        "Intersection size" = (
            intersection_size(
                text = list(size = 0) # text size for count text
            ) +
                labs(x = NULL) +
                theme(
                    axis.text.x = element_blank(),
                    axis.text.y = element_text(size = 8),
                    axis.title.y = element_text(size = 9)
                )
        )
    ),
    
    # Adjust set sizes plot
    set_sizes = (
        upset_set_size() +
            geom_text(
                aes(label = after_stat(count)), # add count labels to show # significant genes in each mu set
                color = "white",
                hjust = -0.1,
                size = 2.5,
                stat = "count"
            ) +
            labs(y = "# significant genes") +
            theme(
                axis.text.y = element_blank(),
                axis.title.y = element_blank(),
                axis.ticks.y = element_blank(),
                
                axis.text.x = element_text(size = 8),
                axis.title.x = element_text(size = 9)
            )
    ),
    
    # Adjust the intersections matrix
    matrix = (
        intersection_matrix(
            geom = geom_point(size = 1.0),
            segment = geom_segment(color = "black", linewidth = 0.25)
        ) +
            #scale_color_nejm() +
            labs(x = NULL, y = NULL)
    ),

    #Define colors for the query sets (affects combination matrix and left bar plot)
    queries = list(
        upset_query(set = "0.001", color = pal_nejm()(1), fill = pal_nejm()(1)),
        upset_query(set = "0.01", color = pal_nejm()(2)[2], fill = pal_nejm()(2)[2]),
        upset_query(set = "0.1", color = pal_nejm()(3)[3], fill = pal_nejm()(3)[3]),
        upset_query(set = "0.5", color = pal_nejm()(4)[4], fill = pal_nejm()(4)[4]),
        upset_query(set = "0.9", color = pal_nejm()(5)[5], fill = pal_nejm()(5)[5]),
        upset_query(set = "0.999", color = pal_nejm()(6)[6], fill = pal_nejm()(6)[6])
    ),
    
    # Use upset_theme to target matrix elements
    themes = list(

        # Intersection Matrix Theme (Dots/Lines); remove x-axis title and text
        intersections_matrix = theme(
            axis.text.x = element_blank(), 
            axis.title.x = element_blank(),
            axis.ticks.x = element_blank(),
            
            axis.text.y = element_text(size = 8)
        )
    )
    
) +

    # Add plot title & remove intersection matrix x lab
    labs(title = "D | Shared significant genes across analyses")


## Save
map(
    .x = c(".png", ".pdf"),
    .f = ~ ggsave(paste0(figures_dir, "R3C2D.mu_sensitivity_gene_upset", .x), 
                  plot = p_r3c2d,
                  width = 6.5, height = 3)
)
```

## Stats

```{r stats}
## Stats: run hypergeometric tests across all combinations
l_sig_genes <- df_x_res_muSENS %>%
    filter(significant == 1) %>%
    group_split(mu) %>%
    map( ~ pull(.x, ensembl_gene_id)) %>%
    set_names(mu_sets)

df_hypergeometric_res <- expand_grid(
    analysis1 = mu_sets,
    analysis2 = mu_sets
) %>% 
    filter(analysis1 != analysis2) %>%
    group_by(grp = paste0(pmin(analysis1, analysis2), "_", pmax(analysis1, analysis2))) %>% 
    slice(1) %>%
    ungroup %>% 
    dplyr::select(-grp) %>% 
    mutate(
        hypergeometric_res = map2(
            .x = .$analysis1,
            .y = .$analysis2,
            .f = ~ f_hypergeometric(gene_list1 = l_sig_genes[[.x]], 
                                    gene_list2 = l_sig_genes[[.y]],
                                    gene_universe = ensembl_gene_ids
            )
        ),
        p_value = map(hypergeometric_res, ~ .x$p_value),
        odds_ratio = map(hypergeometric_res, ~ .x$odds_ratio)
    ) %>% 
    unnest(cols = c(p_value, odds_ratio))

df_hypergeometric_res %>% filter(p_value > 0.05)
df_hypergeometric_res %>% filter(p_value < 0.05)
df_hypergeometric_res %>% 
    mutate(padj = p.adjust(p_value, method = "BH")) %>% 
    filter(padj > 0.05)
```
